# Erie: Relative Abundance Analysis
A relative abundance bar chart helps visualize how the bacterial community abundance levels change over time and season. In our data set, there are three sampling sites: nearshore1, nearshore2, and offshore. Each of the different sites give a partial picture of the overall bacterial makeup of Lake Erie.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=normalizePath(".."))
```

## Setup and reading in data
```{r}
rm(list = ls())
gc()

library(tidyverse)
library(lubridate)

erie <- readRDS("clean_data/erie.rds")
erie_meta <- readRDS("clean_data/erie_meta.rds")
erie_tax <- readRDS("clean_data/erie_tax.rds")
```

## Filter out samples with small read numbers
```{r}
erie_meta <- erie_meta %>%
             filter(sequencingDepth >= 10000)

erie_samples <- erie_meta %>%
                pluck("SampleID")

erie <- erie %>%
        filter(SampleID %>% is.element(erie_samples))
```

## Lengthen the data with a pivot
```{r}
erie <- erie %>%
        select(-c(label, numOtus)) %>%
        pivot_longer(cols = starts_with("Otu"),
                     names_to = "OTU",
                     values_to = "Abundance")
```

## Simplify the taxonomy table
```{r}
erie_tax_phylum <- erie_tax %>%
            select(OTU, Phylum)
```

## Join the erie and erie_tax tables so the OTUs are labeled (Phylum)
```{r}
erie <- erie %>%
        left_join(erie_tax_phylum)
```

## Aggregate the data by SampleID and Phylum

Group the data by SampleID and Phylum
```{r}
erie_phylum <- erie %>%
               group_by(SampleID, Phylum)

rm(erie)
gc()
```

Aggregate Abundance as a sum
```{r}
erie_phylum <- erie_phylum %>%
               summarize(Abundance = sum(Abundance))
```

## Compute relative abundances
```{r}
erie_phylum <- erie_phylum %>%
               mutate(Rel_Abundance = Abundance / sum(Abundance, na.rm = TRUE))
```

## Filter out uncommon phila
How many phyla do we currently have?
```{r}
erie_phylum %>% ungroup() %>% count(Phylum)
```

Removing phyla where the relative abundance is 2% or less
```{r}
erie_phylum <- erie_phylum %>%
               filter(Rel_Abundance > .02)
```

How many phyla do we have after filtering?
```{r}
erie_phylum %>% ungroup() %>% count(Phylum)
```

Saving the filtered dataset
```{r}
erie_phylum %>% ungroup() %>% saveRDS("clean_data/erie_phylum.rds")
```

## Join sample data with the meta data
```{r}
erie_joined <- erie_phylum %>%
               left_join(erie_meta)
```

## Create the stacked bar plots

Defining the set of colors to be used for the stacked bar plot
```{r}
phylum_colors <- c(
  "#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD",
   "#AD6F3B", "#673770","#D14285", "#652926", "#C84248", 
  "#8569D5", "#5E738F","#D1A33D", "#8A7C64", "#599861"
)
```

We will make two plots. The first plot treats the date as a date format, which will highlight the length of time between sample collections. The second plot will treat the dates as characters, removing the irregular spacing to focus on differences between samples
```{r}
ggplot(erie_joined, aes(x = Date, y = Rel_Abundance, fill = Phylum)) + 
  facet_grid(Station~.) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  # Remove x axis title
  theme(axis.title.x = element_blank()) + 
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  ggtitle("Phylum Composition of Lake Erie \n Bacterial Communities by Sampling Site") 
```

```{r}
ggplot(erie_joined, aes(x = format(Date, "%m/%d"), y = Rel_Abundance, fill = Phylum)) + 
  facet_grid(Station~.) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = phylum_colors) +
  scale_x_discrete(
    breaks = c("07/08", "08/04", "09/02", "10/06"),
    labels = c("Jul", "Aug", "Sep", "Oct"),
    drop = FALSE
  ) +
  theme_bw() +
  # Remove x axis title
  theme(axis.title.x = element_blank()) + 
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  ggtitle("Phylum Composition of Lake Erie \n Bacterial Communities by Sampling Site") 
```

## Save the plots
Saving the resulting plots as a pdf
```{r}
ggsave('outputs/Erie_Phylum_relativeAbundance_stackedBar.pdf')
```

## Brief Analysis of the relative abundance plot
### Findings
Comparing across stations, Chlorobi is  more present in late August and early September for the nearshore stations. There are higher levels of Armatimonadetes, Cyanobacteria, and Planctomycetes present in the nearshore stations. Proteobacteria is more common at the offshore station. Finally, Actinobacteria and Bacteroidetes show no significant abundances between stations. Across months, Proteobacteria abundances are the highest in October, and Bacteroides reach their highest offshore concentrations around late August.
