# Erie: Exploratory Data Analysis


## Initial setup and import
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=normalizePath(".."))
```

Clearing the workspace
```{r}
rm(list = ls())
gc()
```

Loading in the required packages for analysis
```{r}
library(tidyverse)
library(vroom)
library(lubridate)
```

Importing the data
```{r}
erie <- readRDS("clean_data/erie.rds")
erie_tax <- readRDS("clean_data/erie_tax.rds")
erie_meta <- readRDS("clean_data/erie_meta.rds")
```

## Creating a histogram of OTU read counts per sample

Aggregate the data by summing the read counts of each OTU column per row
```{r}
erie_sums <- erie %>%
             mutate(reads = rowSums(across(starts_with("Otu"))) %>% as.integer()) %>%
             select(SampleID, reads)
```

Create the histogram of read counts
```{r}
ggplot(data = erie_sums, mapping = aes(x = reads)) +
  geom_histogram(color = "black", fill = "coral", binwidth = 2500) +
  ggtitle("Distribution of sample sequencing depth") +
  xlab("Read counts") +
  theme_bw() +
  theme(axis.title.y = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5))
```

Save the histogram as a pdf
```{r}
ggsave('outputs/Erie_readCounts_Histogram.pdf')
```

## Creating a table summarizing the min, mean, and max read counts
```{r}
erie_sums %>% summarize(min = min(reads), mean = mean(reads), max = max(reads))
```

