# Erie: Exploratory Data Analysis
This file serves as an initial exploration of the Erie dataset. First, we want to look at the sample counts by the number of OTU reads as a histogram. This will give us an idea of the distribution of OTU reads, which tell us how much bacteria was collected in a given sample. More statistical analyses will be conducted in separate documents.

## Initial setup and import
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=normalizePath(".."))
```

Clear the workspace
```{r}
rm(list = ls())
gc()
```

Load in the required packages for analysis
```{r}
library(tidyverse)
library(vroom)
library(lubridate)
library(ggpmisc)
```

Import the data
```{r}
erie <- readRDS("clean_data/erie.rds")
erie_tax <- readRDS("clean_data/erie_tax.rds")
erie_meta <- readRDS("clean_data/erie_meta.rds")

head(erie_meta)
```

## Create a histogram of OTU read counts

Aggregate the data by summing the read counts of each OTU column per row
```{r}
erie_sums <- erie %>%
             mutate(reads = rowSums(across(starts_with("Otu"))) %>% as.integer()) %>%
             select(SampleID, reads)
```

Create the histogram of read counts
```{r}
ggplot(data = erie_sums, mapping = aes(x = reads)) +
  geom_histogram(color = "black", fill = "coral", binwidth = 2500) +
  ggtitle("Distribution of sample sequencing depth") +
  xlab("Read counts") +
  theme_bw() +
  theme(axis.title.y = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5))
```

Save the histogram as a pdf
```{r}
ggsave('outputs/Erie_readCounts_Histogram.pdf')
```

Create a table summarizing the min, mean, and max read counts
```{r}
erie_sums %>% summarize(min = min(reads), mean = mean(reads), max = max(reads))
```


## Create scatter plots to search for correlations
Scatter plot of Temperature vs. Total Dissolved Phosphorus
```{r}
ggplot(data = erie_meta, mapping = aes(x = Temp, y = TDP)) +
  facet_grid(rows = vars(factor(Month,
                                levels = c("June", "July", "August",
                                           "September", "October")))) +
  stat_poly_line(se = FALSE) +
  stat_poly_eq(aes(label = paste(after_stat(eq.label),
                                 after_stat(rr.label),
                                 sep = "*\", \"*"))) +
  geom_point() +
  ggtitle("Temperature vs. Total Dissolved Phosphorus") +
  xlab("Temperature") +
  ylab("TDP") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

Scatter plot of Temperature vs. Turbidity
```{r}
ggplot(data = erie_meta, mapping = aes(x = Turbidity, y = Temp)) +
  facet_grid(rows = vars(factor(Month,
                                levels = c("June", "July", "August",
                                           "September", "October")))) +
  stat_poly_line(se = FALSE) +
  stat_poly_eq(aes(label = paste(after_stat(eq.label),
                                 after_stat(rr.label),
                                 sep = "*\", \"*"))) +
  geom_point() +
  ggtitle("Temperature vs. Turbidity") +
  xlab("Turbidity") +
  ylab("Temperature") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

Scatter plot of Temperature vs. Nitrates
```{r}
ggplot(data = erie_meta, mapping = aes(x = Nitrate, y = Temp)) +
  facet_grid(rows = vars(factor(Month,
                                levels = c("June", "July", "August",
                                           "September", "October")))) +
  stat_poly_line(se = FALSE) +
  stat_poly_eq(aes(label = paste(after_stat(eq.label),
                                 after_stat(rr.label),
                                 sep = "*\", \"*"))) +
  geom_point() +
  ggtitle("Temperature vs. Nitrates") +
  xlab("Nitrates") +
  ylab("Temperature") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

Scatter plot of Nitrates vs. Total Dissolved Phosphorus
```{r}
ggplot(data = erie_meta, mapping = aes(x = TDP, y = Nitrate)) +
  facet_grid(rows = vars(factor(Month,
                                levels = c("June", "July", "August",
                                           "September", "October")))) +
  stat_poly_line(se = FALSE) +
  stat_poly_eq(aes(label = paste(after_stat(eq.label),
                                 after_stat(rr.label),
                                 sep = "*\", \"*"))) +
  geom_point() +
  ggtitle("Nitrates vs. Total Dissolved Phosphorus") +
  xlab("TDP") +
  ylab("Nitrates") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```