# Erie: Cleaning the Data for Analysis

## Initial setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=normalizePath(".."))
```

Clearing the workspace
```{r}
rm(list = ls())
gc()
```

Loading in the required packages for analysis
```{r}
library(tidyverse)
library(vroom)
library(lubridate)
```

## Cleaning the meta file

Loading in the Erie metadata
```{r}
col_types = "cccccccccddddddddddddddddddddddd"
erie_meta <- read_csv("raw_data/habs_metadata_cleaned.csv", col_types = col_types)
```

Converting the Date_year column to the date data type
```{r}
erie_meta <- erie_meta %>%
             mutate(Date = Date_year %>% mdy())
```

Counting the number of samples taken at each station each month
```{r}
erie_meta %>% count(Month, Station)
```

Extract the list of samples from the meta table
```{r}
erie_samples <- erie_meta %>% pluck("SampleID")
```

A summary of the meta data
```{r}
summary(erie_meta[,0:10])
```

## Cleaning the shared file

Load in the shared file column headers and count the number of OTUs
```{r}
shared.path <- "raw_data/chabs.shared"
v1 <- vroom_lines(shared.path, n_max = 1)

countOtuStr <- str_count(v1, pattern = "Otu")
countOtuStr
```

Defining the column types for the tibble
```{r}
col_types <- c("dc", rep("i", countOtuStr)) %>% str_flatten()
```

Load in the shared data file and drop all columns with NA
```{r}
erie <- vroom(shared.path, col_types = col_types) %>%
        select(where(~!all(is.na(.x))))
```

Clean up variables
```{r}
rm(list = c("v1", "countOtuStr", "col_types"))
```

Rename the Group column to SampleID for consistency and filter out all samples that are not present in both the shared and meta files
```{r}
erie <- erie %>%
        rename(SampleID = Group) %>%
        filter(SampleID %>% is.element(erie_samples))
```


```{r}
erie <- erie %>%
        select(-(where(~is.numeric(.x) && sum(.x, na.rm = TRUE) <= 0) & starts_with("Otu")))
```

Extracting the list of OTUs from the shared table
```{r}
erie_otus <- names(erie %>% select(starts_with("Otu")))
```

A summary of the shared data
```{r}
summary(erie[,0:10])
```

## Cleaning the taxonomy file

Loading the taxonomy file
```{r}
erie_tax <- read_tsv("raw_data/chabs-silva.taxonomy")
```

Splitting the taxonomy column into 7 columns
```{r}
tax_levels <- c("Kingdom", "Phylum", "Class", "Order",
                "Family", "Genus", "Species")

erie_tax <- erie_tax %>%
            separate(Taxonomy, tax_levels, sep = ";", extra = "drop") %>%
            mutate(across(all_of(tax_levels), ~str_remove(.x, "(\\(\\d{1,3}\\))$")))
```

Filter out all OTUs that are not present in both the shared and taxonomy files
```{r}
erie_tax <- erie_tax %>%
            filter(OTU %>% is.element(erie_otus))
```

Filtering out taxonomy table for only bacteria in our analyses
```{r}
erie_tax <- erie_tax %>% filter(Kingdom == "Bacteria",
                                Family != "mitochondria",
                                Class != "Chloroplast")
```

Extract the list of OTUs from the taxonomy table
```{r}
erie_otus <- erie_tax %>% pluck("OTU")
```

```{r}
erie <- erie %>%
        select(!starts_with("Otu") | all_of(erie_otus))
```

A summary of the taxonomy data
```{r}
summary(erie_tax)
```

## Writing out the cleaned erie data

```{r}
erie %>% saveRDS("clean_data/erie.rds")
erie_tax %>% saveRDS("clean_data/erie_tax.rds")
erie_meta %>% saveRDS("clean_data/erie_meta.rds")